---
title: "Caliper matrices"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Caliper matrices}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(caliper)
```

## Intro

This vignette explains how to work with Caliper matrices from R using the
`caliper` package. Many of the raw GISDK functions you may be used to have been
streamlined using the `CaliperMatrix` class. The end result is a more
intuitive experience.

This vignette assumes you are already familiar with the basics of the `caliper`
function. If not, see `vignette("using-caliper")`.

Before getting started, I create a temporary copy of this package's matrix file. 
This is only to ensure that changes made by this vignette are not made to the
original file. (This is not necessary in normal workflow.)

```{r}
orig <- system.file(
  "extdata", "gisdk", "testing", "toy_matrix.mtx", package = "caliper"
)
mtx_file <- tempfile(fileext = ".mtx")
ok <- file.copy(orig, mtx_file)
```

## Getting started

Opening a matrix returns a `CaliperMatrix` object with useful attributes and
methods.

```{r}
mtx <- RunFunction("OpenMatrix", mtx_file, NA)
```

Several of R's generic functions can be used on the resulting R6 object.

```{r}
summary(mtx)
```

```{r}
tbl <- as.data.frame(mtx)
head(tbl)
```

The function `as.matrix` can be used on a specific core, or on the matrix object
directly to convert all cores.

```{r}
as.matrix(mtx$core_b)
as.matrix(mtx)
```

You can also send updated data back to your Caliper software. In the code below,
I update the top left corner of `core_b`. Doing so updates the .mtx file.

```{r}
new_data <- matrix(seq(1, 4), nrow = 2, ncol = 2, 
                   dimnames = list(c(1,2), c(1,2)))
new_data
mtx$core_b <- new_data
as.matrix(mtx$core_b)
```

### Full work flow example

The example below introduces one more function: `AddMatrixCore`. Using it, we
create a new core (`core_c`) and fill it with the sum of `core_a` and `core_b`.

```{r}
mtx$AddMatrixCore("core_c")
data <- as.matrix(mtx)
data$core_c <- data$core_a + data$core_b
mtx$core_c <- data$core_c
as.matrix(mtx$core_c)
```

## Working with indices

Row and column indices are how you subset caliper matrices. In the previous
section, we were able to update a subset of matrix cells directly
from R without needing to know about matrix indices; however, knowledge of
indices is valuable when working with more complex GISDK functions.

You can see which indices are availabe in the matrix:

```{r}
mtx$indices
```

The `row_index` and `column_index` active fields allow you to see which index is
active or switch between available indices. In our example matrix, the `subset`
index only contains the first 3 rows and columns of the full matrix.

First, we can check what the currently-active row index is:

```{r}
mtx$row_index
```

The "all" row index is active, which means all 5 rows. We can change it to 
the "subset" index to only work with the first 3.

```{r}
mtx$row_index <- "subset"
as.matrix(mtx$core_b)
```

We can do the same to the column index, which gives us a 3x3 matrix.

```{r}
mtx$column_index <- "subset"
as.matrix(mtx$core_b)
```

Finally, the `caliper` package makes it simple to add new indices. The example
below adds the "new" index and gives the IDs of which rows/columns to include.

```{r}
mtx$AddIndex("new", c(1, 2, 3, 4))
mtx$indices
```

## Handles and currencies

The `caliper` package abstracts away the idea of matrix handles and currencies,
but certain GISDK functions or macros may require them. They can be provided as
shown below.

```{r}
# A GISDK function requiring a matrix handle
RunFunction("GetMatrixBaseIndex", mtx$handle)

# A GISDK function requiring a matrix currency
RunFunction("GetMatrixColumnLabels", mtx$core_a)
```

Note that the column labels are only 1-3 as opposed to 1-5. The matrix currency
pointer respects the `subset` column index we set. One of the best things about
working with Caliper matrices in R is that you never need to create a GISDK
currency. All of that is handled behind the scenes.
