---
title: "Using caliper"
author: "Kyle Ward"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Using caliper}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

`caliper` makes it easy to work with Caliper software from R. Supported
software includes:

  * [TransCAD](https://www.caliper.com/tcovu.htm)
  * [TransModeler](https://www.caliper.com/transmodeler/default.htm)
  * [Maptitude](https://www.caliper.com/maptovu.htm)

The sections below show how to work with the package.

## Connecting

When you load the library, `caliper` will automatically find and connect to
installed Caliper software.

```{r}
library(caliper)
```

It does so using the `connect()` function. If you have multiple
software applications installed, you can use `connect()` to specify which you
want to connect to.

```{r, eval=FALSE}
connect("Maptitude")
```

If you have multiple versions of the same software installed (e.g. TransCAD 7
and TransCAD 8), the last one installed is the one that will be connected to
automatically. To connect to a different version, open that application before
connecting. This can be done manually or from R:

```{r, eval=FALSE}
system2(
  "C:\\Program Files\\TransCAD 9.0\\tcw.exe",
  wait = FALSE, invisible = FALSE
)
connect()
```

To disconnect (and close the software), use:

```{r, eval=FALSE}
disconnect()
```


## Macros and Functions

Caliper software comes with a scripting language called GISDK. In this language,
there is a difference between 'functions' and 'macros'. Functions are written in
C++ and are called directly in GISDK (e.g. `OpenTable()`). Macros are written in
GISDK, and are called using `RunMacro("macro name")`.

### Running a macro

```{r}
RunMacro("G30 Tutorial Folder")
```

```{r, include=FALSE}
folder <- RunMacro("G30 Tutorial Folder")
```

Calling macros from R and GISDK look the same with the `caliper` package.

### Running a function

To run a function from R, the format is slightly different from GISDK. Instead
of calling `OpenTable()` directly, use `RunFunction("OpenTable")`. The example
below opens a FFB file and counts the number of rows in the table.

```{r}
table_name <- RunFunction(
  "OpenTable", "airports", "ffb",
  list(paste0(folder, "airports.bin"))
)
num_rows <- RunFunction("GetRecordCount", table_name, NA)
paste0("The table '", table_name, "' has ", num_rows, " rows.")

# Equivalent in GISDK
# table_name = OpenTable("airports", "ffb", {folder + "airports.bin", null}, )
# num_rows = GetRecordCount(table_name, )
# ShowMessage("The table '" + table_name + "' has " + String(num_rows) + " rows.")
```

The code above introduces two other important differences between R and GISDK:

  * Use a `list` in R when you would use an `array` in GISDK.
  * Use `NA` in R when you would use `null` in GISDK.

To improve integration with R, some GISDK functions have been implemented
directly in the `caliper` package:

  * `RunMacro()`
  * `SetAlternateInterface()`
  * `GetInterface()`
  * `CreateObject()`

Trying to call these functions using `RunFunction()` will display a helpful
error message:

```{r, error=TRUE}
RunFunction("RunMacro")
```

## Custom Macros

In order to call one of your own GISDK functions from R, it must be compiled
into an interface/UI. The `caliper` package comes with a toy GISDK script, which
can be compiled using the `compile_gisdk()` function. Code compiled using
Caliper software also works.

```{r}
rsc_file <- system.file("extdata", "gisdk", "testing", "gisdk.rsc", package = "caliper")
ui_path <- compile_gisdk(rsc_file, tempfile(fileext = ".dbd"))
```

The toy functions are simple, but illustrate some important points in the
examples below.

To reference a custom function that you have written and compiled, first use 
`SetAlternateInterface()` to point to your compiled code. The "first macro"
example simply returns a "Hello World!" message.

```{r}
SetAlternateInterface(ui_path)
RunMacro("first macro")
```

You can see the current interface using

```{r}
GetInterface()
```

and set it back to the default interface with

```{r}
SetAlternateInterface()
GetInterface()
```

## Named Arrays

Many GISDK functions and macros take named arrays (aka options arrays) as input.
`caliper` allows you to pass named lists as an equivalent.

```{r}
SetAlternateInterface(ui_path)

opts <- list()
opts$one <- 1
opts$two <- 2
RunMacro("parse opts array", opts)
```

## Views in R

Caliper software represents tabular data in "views". Earlier examples explained
how to use `RunFunction()` to open a view and get the number of rows using
GISDK functions. It can also be helpful to bring that tabular data into R for
additional processing. Two functions exist to help with this:

  * `view_to_df()`: convert a Caliper view into an R data frame
  * `df_to_view()`: convert a data frame into a Caliper view

In the code below, a data.frame is sent to a view. A new view is created, and
it's name is returned.
  
```{r}
df <- data.frame("one" = c(1, 2), "two" = c(3, 4), "three" = c(5, 6))
df
view <- df_to_view(df)
view
```
  
GISDK functions will now operate on that view as seen in previous examples.

```{r}
RunFunction("GetRecordCount", view, NA)
```

The view can be brought back into a new data.frame.

```{r}
df2 <- view_to_df(view)
df2
```

Finally, the view can be updated with new data. To do so, we include the name
of the view we want to update in our call to `df_to_view()`.

```{r}
df2$one <- 10
view <- df_to_view(df2, view)
view_to_df(view)
```

## Logging and Message Boxes

The behavior of message boxes (errors or otherwise), depends on how Caliper
software is being run from R. Most often, the software is running in the
background (there is no icon on the task bar and no associated window). In this
case, messages are written out to a log file. In the code below, I first write
something to the log and then use `read_log()` to display it.

```{r}
RunFunction("ShowMessage", "This is a test.")
read_log()
```

If a Caliper program is already open when you connect to it, then it will have a
visible window and show up on the task bar. In this case, `ShowMessage` and
related functions will create a window with an OK button. R will wait until you
have closed that pop up before continuing. You can test the behavior by running
the code below. You may have to alt-tab to find the pop up window.

```{r, eval=FALSE}
RunFunction("ShowArray", list(1, 2, 3))
```

## Advanced topics

For more-advanced topics, see the other package vignettes:

  * [Matrices](caliper-matrices.html)
  * [Objects](caliper-objects.html)
  * [Bin Files](bin-files.html)
